A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-09 paper.Rmd
fixing typos in 5.3

## Statistical significance~\label{sig}

@@

## Power comparison of the tests~\label{power-analysis}

@@

## Comparison of test decisions based on $p$-values~\label{p-value}

@@

## Effect of amount of non-linearity~\label{nonlin-analysis}

@@

For visual tests, we expect the "U" shape to be detected more readily, followed by the "S", "M" and "triple-U" shape. From Figure \ref{fig:poly-power-uniform-j}, it can be observed that the power curves mostly align with these expectations, except for the "M" shape, which is as easy to be detected as the "S" shape. This suggests a benefit of the visual test, the knowing the shape ahead of time is not needed for its application.

@@

## Effect of shape of heteroskedasticity~\label{hetero-analysis}

>>>>

## Statistical significance\label{sig}

@@

## Power comparison of the tests\label{power-analysis}

@@

## Comparison of test decisions based on $p$-values\label{p-value}

@@

## Effect of amount of non-linearity\label{nonlin-analysis}

@@

For visual tests, we expect the "U" shape to be detected more readily, followed by the "S", "M" and "triple-U" shape. From Figure \ref{fig:poly-power-uniform-j}, it can be observed that the power curves mostly align with these expectations, except for the "M" shape, which is as easy to be detected as the "S" shape. This suggests a benefit of the visual test: knowing the shape ahead of time is *not* needed for its application.

@@

## Effect of shape of heteroskedasticity\label{hetero-analysis}

### 2023-04-09 paper.tex
fixing spelling in 5.2
fixing typos in 5.3

We can assess residuals in multiple ways. Residuals may be plotted as a
histogram or quantile-quantile plot to examine the distribution. Using
the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, we consider it to be well
behaved. However, if the distribution is skewed, bimodal, multimodal, or
contains outliers, there is a cause for concern. One could also inspect
the distribution by conducting a goodness of fit test, such as the
Shapiro-Wilk Normality test \citep{shapiro1965analysis}.

More typically, residuals will be plotted, as a scatter plot against the
predicted values and each of the explanatory variables to scrutinize
their relationships. If there are any visually discoverable patterns,
the model is potentially misspecified. In general, one looks for
noticeable departures from the model like non-linear dependency or
heteroskedasticity. However, correctly judging a residual plot where no
pattern exists can be a painstakingly difficult task for humans. It is
especially common, particularly among students, to misinterpret patterns
that are random noise and random deviation from a model
\citep{loy2021bringing}. It is also possible to conduct hypothesis tests
for non-linear dependence \citep{ramsey_tests_1969}, and use a
Breusch-Pagan test \citep{breusch_simple_1979} for heteroskedasticity.

@@

Many different hypothesis tests are available to detect specific model
defects. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the Breusch-Pagan
test \citep{breusch_simple_1979}, which are both derived from the
Lagrange multiplier test \citep{silvey1959lagrangian} principle that
relies on the asymptotic properties of the null distribution. To test
specific forms of non-linearity, one may apply the F-test as a model
structural test to examine the significance of specific polynomial and
non-linear forms of the predictors, or the significance of proxy
variables as in the Ramsey Regression Equation Specification Error Test
(RESET) \citep{ramsey_tests_1969}. The Shapiro-Wilk test
\citep{shapiro1965analysis} is the most widely used test of
non-normality included by many of the statistical software programs. The
Jarque--Bera test \citep{jarque1980efficient} is also used to directly
check whether the sample skewness and kurtosis match a normal
distribution.

Table \ref{tab:example-residual-plot-table} displays the \(p\)-values
from the RESET, Breusch-Pagan and Shapiro-Wilk tests applied to the
residual plots in Figure \ref{fig:residual-plot-common-departures}. The
RESET test and Breusch-Pagan test were computed using the
\texttt{resettest} and \texttt{bptest} functions from the R package
\texttt{lmtest}, respectively. The Shapiro-Wilk test was computed using
the \texttt{shapiro.test} from the core R package \texttt{stats}. (The
the R package \texttt{skedastic} \citep{skedastic} contains a large
collection of tests for heteroskedasticity.) Although, the RESET test is
exact, it requires the selection of a power parameter. According to
\citet{ramsey_tests_1969}, a power of four is recommended, which we
adopted in our analysis. The Breusch-Pagan and Shapiro-Wilk tests are
approximate.

We would expect the RESET test for non-linearity to reject residual plot
B, the Breusch-Pagan test for heteroskedasticity to reject the residual
plot C, and Shapiro-Wilk test for non-normality to reject residual plot
D, which they all do and correctly fail to reject residual plot A.
Interestingly, the Breusch-Pagan and Shapiro-Wilk tests also reject the
residual plots exhibiting structure that they weren't designed for.
\citet{cook1982residuals} discusses that most residual-based tests for
particular types of departure from model assumptions are also sensitive
to other types of departures, that is, correctly rejected but for the
wrong reason, a phenomenon known as the ``Type III error''. Also, some
types of departure can also have elements of other types of departure,
for example, non-linearity could be viewed as heteroskedasticity.
Additionally, other problesms such as outliers can trigger the rejection
\citep{cook_applied_1999}.

@@

\caption{\label{tab:example-residual-plot-table}Statistical significance testing for departures from good residuals for plots in Figure \ref{fig:residual-plot-common-departures}. Shown are the $p$-values calculated for the RESET, the Breusch-Pagan and the Shapiro–Wilk tests. The good residual plot (A) is judged a good residual plot, as expected, by all tests. The non-linearity (B) is detected by all tests, as might be expected given the extreme structure.}
\centering
\begin{tabular}[t]{llrrr}
\toprule
Plot & Departures & RESET & Breusch-Pagan & Shapiro–Wilk\\
\midrule
A & None & 0.779 & 0.133 & 0.728\\
B & Non-linearity & \em{0.000} & \em{0.000} & \em{0.039}\\
C & Heteroskedasticity & 0.658 & \em{0.000} & \em{0.000}\\
D & Non-normality & 0.863 & 0.736 & \em{0.000}\\
\bottomrule
\end{tabular}
\end{table}

@@

However, it is practically impossible to test this specific \(H_0\) with
conventional tests, which are constructed to measure specific
departures. For example, the RESET test is formulated as
\(H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0\) against
\(H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0\),
from
\(y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)\).
Similarly, the Breusch-Pagan test is designed to specifically test
\(H_0:\) \emph{error variances are all equal}
(\(\zeta_i=0 \text{ for } i=1,..,p\)) versus the alternative \(H_a:\)
\emph{that the error variances are a multiplicative function of one or
more variables} (\(\text{at least one } \zeta_i\neq 0\)) from
\(e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)\).

@@

In terms of lineups containing heteroskedasticity patterns, 76\% are
rejected by conventional tests, while 56\% are rejected by visual tests.
When the conventional test rejects a lineup, there is a good chance
(73\%) the visual test will also reject it.

Surprisingly, the visual test rejects 1 of the 33 (3\%) of lineups where
the conventional test does not reject. Figure \ref{fig:heter-example}
shows this lineup. The data plot in position seventeen displays a
relatively strong heteroskedasticity pattern, and has a strong effect
size (\(log_e(E)=4.02\)). This is reflected by the visual test
\(p\text{-value} = 0.026\). But the Breusch-Pagan test
\(p\text{-value} = 0.056\), is slightly above the significance cutoff of
\(0.05\). This lineup was evaluated by 11 subjects, it has experimental
factors \(a = 0\) (``butterfly'' shape), \(b = 64\) (large variance
ratio), \(n = 50\) (small sample size), and a uniform distribution for
the fitted values. It may have been the small sample size and the
presence of a few outliers that may have resulted in the lack of
detection by the conventional test.

@@

For visual tests, we expect the ``U'' shape to be detected more readily,
followed by the ``S'', ``M'' and ``triple-U'' shape. From Figure
\ref{fig:poly-power-uniform-j}, it can be observed that the power curves
mostly align with these expectations, except for the ``M'' shape, which
is as easy to be detected as the ``S'' shape. This suggests a benefit of
the visual test, the knowing the shape ahead of time is not needed for
its application.

@@

\caption{Comparison of power on lineups with different fitted value distributions for conventional and visual tests for non-linearity and heteroskedasticity patterns. The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and Breusch-Pagan tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power.}\label{fig:different-x-dist-poly-power}
\end{figure}

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the
fitted value distribution on the power of conventional (top row) and
visual (bottom row) tests for both the non-linearity (left column) and
heteroskedasticity (right column) patterns. For conventional tests, only
the power curves of appropriate tests which are RESET tests and
Breusch-Pagan tests for non-linearity pattern and heteroskedasticity
pattern respectively are shown. For visual tests, note that, more
evaluations on lineup with uniform fitted value distribution were
collected, so to have a fair comparison, we randomly sample five from
the total eleven evaluations to estimate the power curves. Re-sampling
produces the multiple curves for the uniform, and provides an indication
of the variation of the power.

>>>>

We can assess residuals in multiple ways. Residuals may be plotted as a
histogram or quantile-quantile plot to examine the distribution. Using
the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, we consider it to be well
behaved. However, if the distribution is skewed, bimodal, multimodal, or
contains outliers, there is a cause for concern. One could also inspect
the distribution by conducting a goodness of fit test, such as the
Shapiro-Wilk (SW) Normality test \citep{shapiro1965analysis}.

More typically, residuals will be plotted, as a scatter plot against the
predicted values and each of the explanatory variables to scrutinize
their relationships. If there are any visually discoverable patterns,
the model is potentially misspecified. In general, one looks for
noticeable departures from the model like non-linear dependency or
heteroskedasticity. However, correctly judging a residual plot where no
pattern exists can be a painstakingly difficult task for humans. It is
especially common, particularly among students, to misinterpret patterns
that are random noise and random deviation from a model
\citep{loy2021bringing}. It is also possible to conduct hypothesis tests
for non-linear dependence \citep{ramsey_tests_1969}, and use a
Breusch-Pagan (BP) test \citep{breusch_simple_1979} for
heteroskedasticity.

@@

Many different hypothesis tests are available to detect specific model
defects. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the BP test
\citep{breusch_simple_1979}, which are both derived from the Lagrange
multiplier test \citep{silvey1959lagrangian} principle that relies on
the asymptotic properties of the null distribution. To test specific
forms of non-linearity, one may apply the F-test as a model structural
test to examine the significance of specific polynomial and non-linear
forms of the predictors, or the significance of proxy variables as in
the Ramsey Regression Equation Specification Error Test (RESET)
\citep{ramsey_tests_1969}. The SW test \citep{shapiro1965analysis} is
the most widely used test of non-normality included by many of the
statistical software programs. The Jarque--Bera test
\citep{jarque1980efficient} is also used to directly check whether the
sample skewness and kurtosis match a normal distribution.

Table \ref{tab:example-residual-plot-table} displays the \(p\)-values
from the RESET, BP and SW tests applied to the residual plots in Figure
\ref{fig:residual-plot-common-departures}. The RESET test and BP test
were computed using the \texttt{resettest} and \texttt{bptest} functions
from the R package \texttt{lmtest}, respectively. The SW test was
computed using the \texttt{shapiro.test} from the core R package
\texttt{stats}. (The the R package \texttt{skedastic} \citep{skedastic}
contains a large collection of tests for heteroskedasticity.) Although,
the RESET test is exact, it requires the selection of a power parameter.
According to \citet{ramsey_tests_1969}, a power of four is recommended,
which we adopted in our analysis. The BP and SW tests are approximate.

We would expect the RESET test for non-linearity to reject residual plot
B, the BP test for heteroskedasticity to reject the residual plot C, and
SW test for non-normality to reject residual plot D, which they all do
and correctly fail to reject residual plot A. Interestingly, the BP and
SW tests also reject the residual plots exhibiting structure that they
weren't designed for. \citet{cook1982residuals} discusses that most
residual-based tests for particular types of departure from model
assumptions are also sensitive to other types of departures, that is,
correctly rejected but for the wrong reason, a phenomenon known as the
``Type III error''. Also, some types of departure can also have elements
of other types of departure, for example, non-linearity could be viewed
as heteroskedasticity. Additionally, other problesms such as outliers
can trigger the rejection \citep{cook_applied_1999}.

@@

\caption{\label{tab:example-residual-plot-table}Statistical significance testing for departures from good residuals for plots in Figure \ref{fig:residual-plot-common-departures}. Shown are the $p$-values calculated for the RESET, the BP and the SW tests. The good residual plot (A) is judged a good residual plot, as expected, by all tests. The non-linearity (B) is detected by all tests, as might be expected given the extreme structure.}
\centering
\begin{tabular}[t]{llrrr}
\toprule
Plot & Departures & RESET & BP & SW\\
\midrule
A & None & 0.779 & 0.133 & 0.728\\
B & Non-linearity & \em{0.000} & \em{0.000} & \em{0.039}\\
C & Heteroskedasticity & 0.658 & \em{0.000} & \em{0.000}\\
D & Non-normality & 0.863 & 0.736 & \em{0.000}\\
\bottomrule
\end{tabular}
\end{table}

@@

However, it is practically impossible to test this specific \(H_0\) with
conventional tests, which are constructed to measure specific
departures. For example, the RESET test is formulated as
\(H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0\) against
\(H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0\),
from
\(y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)\).
Similarly, the BP test is designed to specifically test \(H_0:\)
\emph{error variances are all equal}
(\(\zeta_i=0 \text{ for } i=1,..,p\)) versus the alternative \(H_a:\)
\emph{that the error variances are a multiplicative function of one or
more variables} (\(\text{at least one } \zeta_i\neq 0\)) from
\(e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)\).

@@

In terms of lineups containing heteroskedasticity patterns, 76\% are
rejected by conventional tests, while 56\% are rejected by visual tests.
The visual test rejects 73\% of the lineups that the conventional visual
test reject.

Surprisingly, the visual test rejects 1 of the 33 (3\%) of lineups where
the conventional test does not reject. Figure \ref{fig:heter-example}
shows this lineup. The data plot in position seventeen displays a
relatively strong heteroskedasticity pattern, and has a strong effect
size (\(log_e(E)=4.02\)). This is reflected by the visual test
\(p\text{-value} = 0.026\). But the BP test \(p\text{-value} = 0.056\),
is slightly above the significance cutoff of \(0.05\). This lineup was
evaluated by 11 subjects, it has experimental factors \(a = 0\)
(``butterfly'' shape), \(b = 64\) (large variance ratio), \(n = 50\)
(small sample size), and a uniform distribution for the fitted values.
It may have been the small sample size and the presence of a few
outliers that may have resulted in the lack of detection by the
conventional test.

@@

For visual tests, we expect the ``U'' shape to be detected more readily,
followed by the ``S'', ``M'' and ``triple-U'' shape. From Figure
\ref{fig:poly-power-uniform-j}, it can be observed that the power curves
mostly align with these expectations, except for the ``M'' shape, which
is as easy to be detected as the ``S'' shape. This suggests a benefit of
the visual test: knowing the shape ahead of time is \emph{not} needed
for its application.

@@

\caption{Comparison of power on lineups with different fitted value distributions for conventional and visual tests for non-linearity and heteroskedasticity patterns. The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power.}\label{fig:different-x-dist-poly-power}
\end{figure}

Figure \ref{fig:different-x-dist-poly-power} examines the impact of the
fitted value distribution on the power of conventional (top row) and
visual (bottom row) tests for both the non-linearity (left column) and
heteroskedasticity (right column) patterns. For conventional tests, only
the power curves of appropriate tests which are RESET tests and BP tests
for non-linearity pattern and heteroskedasticity pattern respectively
are shown. For visual tests, note that, more evaluations on lineup with
uniform fitted value distribution were collected, so to have a fair
comparison, we randomly sample five from the total eleven evaluations to
estimate the power curves. Re-sampling produces the multiple curves for
the uniform, and provides an indication of the variation of the power.
